### 2022-09-19 chapter2/chapter.tex
finito

In this chapter we introduce the discontinuous Galerkin (DG) method applied to fluid queues to approximate the operator-analytic expressions for fluid-fluid queues in \cite{bo2014}. 
\begin{center}
	\begin{minipage}{0.8\textwidth}
		\textit{Apart from Sections~\ref{sec: other applications} and \ref{sec: limiting and linearity}, this chapter has been taken from Section~4 and 5 of \cite{blnos2022} with only minor changes, such as notation, so that this chapter is consistent with the rest of the thesis. I am a co-author of the paper \cite{blnos2022}. %The conceptualisation of \cite{blnos2022} was originally by Vikram Sunkara, Nigel Bean and Giang Nguyen, and the original coding was done by Vikram Sunkara. I made significant contributions to Section~3 of the paper, expressing the operator-theoretic expressions to use the same partition as the approximation scheme. I contributed Sections~4.4 and 5.1. I extended the numerical experiments in Section~6 to higher orders and made all the plots in Section~6. Appendix~A is also my original work. I did a significant proportion of the writing of the manuscript and addressed the reviewers comments and also developed code for the numerical experiments.
		}
	\end{minipage}
	\end{center}
% The DG method for B 
    % direct copy paste from the paper
% Problems with DG, non-neg/oscillations
    % solutions: filtering and limiting
% Approximating R: projection, interpolation, cell averages (eluding to QBD-RAP and unif method)
% Approximating Psi (and the limiting distribution, but this isnt really necessary)
    % constructing D
    % solving the matrix-Riccati equation, iteration/algorithm.

@@

\section{The partial differential equation}
We start by introducing the PDE (assuming the PDE exists) from which we will extract the approximation to the generator \(\mathbb B\).

@@

The DG method proceeds by first considering the \textit{weak-formulation} of the PDE, which is constructed from the \textit{strong-form} of the PDE, Equation~(\ref{eqn:pde_density_matrix}). In general, to construct the weak-form we need a set of test functions, say \(W\). Now, take the strong form of the PDE, multiply it by some test function \(\psi(x)\in W\), integrate with respect to \(x\), and apply integration by parts to get 
\begin{align}
\begin{multlined}[t]
	\int_{x\in\mathbb R}\cfrac{\partial}{\partial t} f_j(x,t)\psi(x)\wrt x = \int_{x\in\mathbb R} \sum_{i\in\calS}f_i(x,t)T_{ij}\psi(x)\wrt x 
	%
	+  \int_{x\in\mathbb R} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\psi(x)\wrt x \\{}- [ f_j(x,t)c_j\psi(x)]_{x=-\infty}^{x=\infty}, \end{multlined}\label{eqn:weak form}
\end{align}
for \(j\in\calS\). It is convenient to choose \(W\) such that \(\psi(-\infty)=\psi(\infty)=0\), in which case the last term on the right is zero. Requiring (\ref{eqn:weak form}) to hold for every \(\psi\in W\) gives the weak-formulation of the PDE. Solutions to (\ref{eqn:weak form}) are known as \textit{weak} solutions and generalise the concept of a solution of the PDE. For example, this may allow discontinuities with respect to \(x\) in the solution -- something which is ill-defined for the strong form.

@@

It remains to approximate the \textit{flux}, \(f_j(x,t)\) at the cell edges \(y_k,\,k\in\mathbb Z\), so that we may evaluate the terms \([f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}\), \(r=1,...,p_k,\,k\in\mathbb Z\). This is the key for DG -- it joins the local approximations on each cell into a global approximation on the whole domain of approximation. The flux is the instantaneous rate (with respect to time) at which density moves across the boundaries \(y_k,\,k\in\mathbb Z\). There are different choices for the flux, and we refer the reader to \citep{c99,nodalDGBook}, and references therein, for some discussion of the topic. Here, we choose the \textit{upwind} scheme, which, as we shall see, closely resembles the flux terms from the generator \(\mathbb B\). The approximate flux, also known as the \textit{numerical flux}, is given by 
\[f^*_j(x,t) = sign(c_j)\lim_{\varepsilon\to0^+}u_j(x-\varepsilon c_j,t),\]
at each \(x=y_k,\,k\in\mathbb Z\). 
Intuitively, the upwind flux takes the value of the density immediately on the upwind side of each \(y_k\).

@@

Indeed, the DG scheme has the potential to answer many mathematical questions for fluid queues. However, one reason DG schemes may not have found such widespread use in the context of fluid queues is the difficulties surrounding discontinuous solutions, which we elaborate on in Section~\ref{sec: slope limiting} and demonstrate in Chapter~\ref{sec: numerics} (see also Section~5.6~\cite{nodalDGBook}), which we often want to consider. Moreover, there are sometimes other techniques which can give exact solutions, for example exact expressions for transient quantities are derived in \cite{rs2003,ar2004,bean2005,bean2009}, and for the limiting distribution are derived in \cite{ajr2005,dasilva2005,lnp13,s2017}.

Nonetheless, the DG scheme can be used to answer a very broad range of questions about fluid queues (such as transient distributions, hitting times on complex geometries, and of course analysis of fluid-fluid queues) which have no exact, readily computable solution. Moreover, we will numerically investigate the performance of the DG scheme (and other numerical schemes which we will develop) in Chapter~\ref{sec: numerics} which require numerical integration with respect to time. Hence, we now briefly introduce the numerical time integration scheme which we use in this thesis, as well as some issues pertaining to oscillatory approximations.

% APPLICATION TO INTERGRATING OVER TIME
\section{Remarks on slope limiters, linear operators and application to fluid-fluid queues}\label{sec: limiting and linearity}
The process of slope limiting, as described in Section~\ref{sec: slope limiting}, essentially amounts to a non-linear post-processing of the approximate solution after applying the operator \(\bs B\). Thus, the slope-limited scheme arrives at an approximation which is the action of a non-linear operator applied to the initial condition. The fact that the approximate operator is non-linear makes solving the Riccati Equation~(\ref{eqn:RiccatiPsi}) difficult/impossible, whereas there are efficient techniques to solve matrix-Riccati equations \cite{bot08}. Thus, for the application to fluid-fluid queues slope limiting is not a viable option. We could, however, proceed with a vanilla DG scheme (with no limiter) and, when we apply the operator \(\bs \Psi\) to some initial condition and realise an approximation to the distribution of the fluid queue when \(\{Y(t)\}\) first returns to its initial level, we could then post-process the solution with a limiter or filter if oscillations are present in the approximation. The application of the limiter/filter is dependent on the initial condition, so the post-processing needs to be done for every initial condition under consideration. If there are many initial distributions to consider, this could result in an increased computational cost. We do not investigate this post-processing approach in this thesis.

On a related note, the advantage of a linear approximation can also be useful if we want to approximate transient distributions of fluid queues for many initial conditions. For example, consider solving (\ref{eqn: DG ODE w BCs 3}) numerically with a forward-Euler scheme to integrate over time. To advance the coefficients from time \(t\) to time \(t+h\) we compute
\begin{align*}
	&\bs a(t+h) = \bs a(t) + h \bs a(t)\bs Q = \bs a(t)\left[\bs I + h\bs Q\right].
\end{align*} 
Thus, the coefficients at time \(t_0=nh\) are given by 
\begin{align*}
	&\bs a(nh) = \bs a(0)\left[\bs I + h\bs Q\right]^n,
\end{align*} 
which is a linear transformation of the coefficients \(\bs a(0)\). Thus, if we determine the matrix \(\left[\bs I + h\bs Q\right]^n\), then we may apply it to any initial condition to approximate the solution at time \(t_0=nh\). In the presence of a limiter, this is not possible as the coefficients at time \(t_0=nh\) are not a linear transformation of the initial coefficients.

Another point to note is that limiters incur a computational cost. There is a computational cost in reducing the solution to linear when oscillations are detected, and also in determining whether there are oscillations or not. Thus, even if the solution needs no limiting, there is an added computational cost in determining this is the case. Given the limiter is applied for each stage and each time-step of the time integration scheme in Section~\ref{subsec: slope lim and int}, then this computational cost may not be insignificant.

One advantage of the QBD-RAP (and uniformisation) scheme we develop in this thesis is that the resulting operator is positivity-preserving and linear.
% This raises two points. First, the concept of limiting applies only to problems with the concept of an approximate solution, and not to problems where the object of interest is the approximate operator itself. For example, we solve the Ricatti equation~\ref{eqn:RiccatiPsi} for the matrix \(\bs \Psi\) to find an approximation to \(\mathbb \Psi\). There is no concept of an approximate solution in this problem, only the concepts of the approximate operators. Thus, there is no concept of slope limiting
% 
% added computational cost when lots of limiting is needed
%

>>>>

In this chapter we introduce the discontinuous Galerkin (DG) method applied to fluid queues to approximate the operator-analytic expressions for fluid-fluid queues in \cite{bo2014}. 
\begin{center}
	\begin{minipage}{0.8\textwidth}
		\textit{Apart from Section~\ref{sec: other applications}, this chapter has been taken from Section~4 and 5 of \cite{blnos2022} with only minor changes, such as notation, so that this chapter is consistent with the rest of the thesis. I am a co-author of the paper \cite{blnos2022}. %The conceptualisation of \cite{blnos2022} was originally by Vikram Sunkara, Nigel Bean and Giang Nguyen, and the original coding was done by Vikram Sunkara. I made significant contributions to Section~3 of the paper, expressing the operator-theoretic expressions to use the same partition as the approximation scheme. I contributed Sections~4.4 and 5.1. I extended the numerical experiments in Section~6 to higher orders and made all the plots in Section~6. Appendix~A is also my original work. I did a significant proportion of the writing of the manuscript and addressed the reviewers comments and also developed code for the numerical experiments.
		}
	\end{minipage}
	\end{center}
% The DG method for B 
    % direct copy paste from the paper
% Problems with DG, non-neg/oscillations
    % solutions: filtering and limiting
% Approximating R: projection, interpolation, cell averages (eluding to QBD-RAP and unif method)
% Approximating Psi (and the limiting distribution, but this isnt really necessary)
    % constructing D
    % solving the matrix-Riccati equation, iteration/algorithm.

@@

\section{The partial differential equation}
We start by recalling the PDE (assuming the PDE exists) from which we will extract the approximation to the generator \(\mathbb B\).

@@

The DG method proceeds by first considering the \textit{weak-formulation} of the PDE, which is constructed from the \textit{strong-form} of the PDE, Equation~(\ref{eqn:pde_density_matrix}). In general, to construct the weak-form we need a set of test functions, say \(W\). Now, take the strong form of the PDE, multiply it by some test function \(\psi(x)\in W\), integrate with respect to \(x\), and apply integration by parts to get 
\begin{align}
\begin{multlined}[t]
	\int_{x\in\mathbb R}\cfrac{\partial}{\partial t} f_j(x,t)\psi(x)\wrt x = \int_{x\in\mathbb R} \sum_{i\in\calS}f_i(x,t)T_{ij}\psi(x)\wrt x 
	%
	+  \int_{x\in\mathbb R} f_j(x,t)c_j\cfrac{\wrt}{\wrt x}\psi(x)\wrt x \\{}- [ f_j(x,t)c_j\psi(x)]_{x=-\infty}^{x=\infty}, \end{multlined}\label{eqn:weak form}
\end{align}
for \(j\in\calS\). It is convenient to choose \(W\) such that \(\psi(-\infty)=\psi(\infty)=0\), in which case the last term on the right is zero. %Requiring (\ref{eqn:weak form}) to hold for every \(\psi\in W\) gives the weak-formulation of the PDE. Solutions to (\ref{eqn:weak form}) are known as \textit{weak} solutions and generalise the concept of a solution of the PDE. %For example, this may allow discontinuities with respect to \(x\) in the solution -- something which is ill-defined for the strong form.

@@

It remains to approximate the \textit{flux}, \(f_j(x,t)\) at the cell edges \(y_k,\,k\in\mathbb Z\), so that we may evaluate the terms \([f_j(x,t)\phi^r_k(x)]_{x=y_k}^{x=y_{k+1}}\), \(r=1,...,p_k,\,k\in\mathbb Z\). %This is the key for DG -- it joins the local approximations on each cell into a global approximation on the whole domain of approximation. 
The flux is the instantaneous rate (with respect to time) at which density moves across the boundaries \(y_k,\,k\in\mathbb Z\). There are different choices for the flux, and we refer the reader to \citep{c99,nodalDGBook}, and references therein, for some discussion of the topic. Here, we choose the \textit{upwind} scheme, which, as we shall see, closely resembles the flux terms from the generator \(\mathbb B\). The approximate flux, also known as the \textit{numerical flux}, is given by 
\[f^*_j(x,t) = sign(c_j)\lim_{\varepsilon\to0^+}u_j(x-\varepsilon c_j,t),\]
at each \(x=y_k,\,k\in\mathbb Z\). 
Intuitively, the upwind flux takes the value of the density immediately on the upwind side of each \(y_k\).

@@

Indeed, the DG scheme has the potential to answer many mathematical questions for fluid queues. However, one reason DG schemes may not have found such widespread use in the context of fluid queues is the difficulties surrounding discontinuous solutions, which is something we often want to consider. Moreover, there are sometimes other techniques which can give exact solutions, for example exact expressions for transient quantities are derived in \cite{rs2003,ar2004,bean2005,bean2009}, and for the limiting distribution are derived in \cite{ajr2005,dasilva2005,lnp13,s2017}. Nonetheless, the DG scheme can be used to answer a very broad range of questions about fluid queues (such as transient distributions, hitting times on complex geometries, and of course analysis of fluid-fluid queues) which have no exact, readily computable solution.% Moreover, we will numerically investigate the performance of the DG scheme (and other numerical schemes which we will develop) in Chapter~\ref{sec: numerics} which require numerical integration with respect to time. Hence, we now briefly introduce the numerical time integration scheme which we use in this thesis, as well as some issues pertaining to oscillatory approximations.

% APPLICATION TO INTERGRATING OVER TIME
% \section{Remarks on slope limiters, linear operators and application to fluid-fluid queues}\label{sec: limiting and linearity}
% The process of slope limiting, as described in Section~\ref{sec: slope limiting}, essentially amounts to a non-linear post-processing of the approximate solution after applying the operator \(\bs B\). Thus, the slope-limited scheme arrives at an approximation which is the action of a non-linear operator applied to the initial condition. The fact that the approximate operator is non-linear makes solving the Riccati Equation~(\ref{eqn:RiccatiPsi}) difficult/impossible, whereas there are efficient techniques to solve matrix-Riccati equations \cite{bot08}. Thus, for the application to fluid-fluid queues, slope limiting is not a viable option. %We could, however, proceed with a vanilla DG scheme (with no limiter) and, when we apply the operator \(\bs \Psi\) to some initial condition and realise an approximation to the distribution of the fluid queue when \(\{Y(t)\}\) first returns to its initial level, we could then post-process the solution with a limiter or filter if oscillations are present in the approximation. The application of the limiter/filter is dependent on the initial condition, so the post-processing needs to be done for every initial condition under consideration. If there are many initial distributions to consider, this could result in an increased computational cost. We do not investigate this post-processing approach in this thesis.

% On a related note, a linear-operator can also be useful if we want to approximate transient distributions of fluid queues for many initial conditions. For example, consider solving (\ref{eqn: DG ODE w BCs 3}) numerically with a forward-Euler scheme to integrate over time. To advance the coefficients from time \(t\) to time \(t+h\) we compute
% \begin{align*}
% 	&\bs a(t+h) = \bs a(t) + h \bs a(t)\bs Q = \bs a(t)\left[\bs I + h\bs Q\right].
% \end{align*} 
% Thus, the coefficients at time \(t_0=nh\) are given by 
% \begin{align*}
% 	&\bs a(nh) = \bs a(0)\left[\bs I + h\bs Q\right]^n,
% \end{align*} 
% which is a linear transformation of the coefficients \(\bs a(0)\). Thus, if we determine the matrix \(\left[\bs I + h\bs Q\right]^n\), then we may apply it to any initial condition to approximate the solution at time \(t_0=nh\). In the presence of a limiter, this is not possible as the coefficients at time \(t_0=nh\) are not a linear transformation of the initial coefficients.

% Another point to note is that limiters incur a computational cost. There is a computational cost in reducing the solution to linear when oscillations are detected, and also in determining whether there are oscillations or not. Thus, even if the solution needs no limiting, there is an added computational cost in determining this is the case. Given the limiter is applied for each stage and each time-step of the time integration scheme in Section~\ref{subsec: slope lim and int}, then this computational cost may not be insignificant.

% One advantage of the QBD-RAP (and uniformisation) scheme we develop in this thesis is that the resulting operator is positivity-preserving and linear.
% This raises two points. First, the concept of limiting applies only to problems with the concept of an approximate solution, and not to problems where the object of interest is the approximate operator itself. For example, we solve the Ricatti equation~\ref{eqn:RiccatiPsi} for the matrix \(\bs \Psi\) to find an approximation to \(\mathbb \Psi\). There is no concept of an approximate solution in this problem, only the concepts of the approximate operators. Thus, there is no concept of slope limiting
% 
% added computational cost when lots of limiting is needed
%

### 2022-09-19 chapter6/chapter.tex
finito

We implement two slope limited DG schemes, the DG-lim and DG-lin-lim scheme, which we describe in more detail in the next section. The numerical experiments demonstrate that the DG-lim scheme does not lead to a viable approximation scheme in the presence of discontinuities, while the DG-lin-lim scheme results in a relatively performant approximation scheme for the discontinuous problems considered here. In general, the QBD-RAP scheme performs similarly to the DG-lin-lim scheme, while the uniformisation scheme is a viable approximation method, but produces larger errors than the QBD-RAP and DG-lin-lim scheme, for the examples considered here.

Furthermore, when slope limiting is used, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter (or filter), which could be significant, particularly if many initial conditions are to be considered. In contrast, for the uniformisation and QBD-RAP, the approximate solution is guaranteed to produce positive probabilities without any post-processing or extra computation.

@@

The concept of slope-limiting originates in the context of numerically integrating (typically with respect to a \emph{time} variable) a so-called \emph{semi-discrete system}, as we do in Section~\ref{sec: transient approx}. However, approximating the operator \(\mathbb\Psi\) of a fluid-fluid queue amounts to solving an implicit matrix-equation and there is no concept of numerical integration with respect to time. Hence, the concept of slope-limiting, it is not directly transferable. One could process the initial condition and post-process the solution with a limiter. However, there are possibly better choices for this post-processing, such as filtering, \citep[Section~5.6.1]{nodalDGBook}. One could also view the composition of the DG operator and the limiting operator as one, but the result is a non-linear operator which is of no use for simplifying the Riccati Equation which characterises \(\mathbb \Psi\). Considering there is no theory on the use of the DG scheme to approximate \(\mathbb \Psi\) and hence the resultant approximation is not rigorously-defined, the author does not know if the use of a limiter in this context is possible.

@@

We obtain approximations to the generator of the fluid queue \(\{(X(t),\varphi(t))\}\) via the DG, QBD-RAP and uniformisation schemes (as discussed at the start of this chapter, we do not apply slope limiting for this calculation). All the discretisations use a cell width of \(\Delta=0.4\). The approximate generator matrix is then used to approximate the first-return operator \(\mathbb\Psi(s)\) as discussed in Section~\ref{sec: intro Psi}. As in Section~\ref{sec: intro Psi}, let \(\bs{B}^{m,n}\), \(m,n\in\{+,-,0\}\) be an approximation to \(\mathbb B^{m,n},\, m,n\in\{+,-,0\}\) of dimension \(p\), and recall that \(\bs{D}^{n,m},\,m,n\in\{+,-\}\) is given by 
\[\bs D^{m,n} = \bs{R}^m (\bs{B}^{m,n} + \bs{B}^{m,0}(\bs B^{0,0})^{-1}\bs{B}^{0,n}), \, m,n\in\{+,-\}, \]
where \(\bs R^m\) is a diagonal matrix with diagonal blocks \(1/r_i^k\bs{I}_p\), \(i\in\calS_k^m\), \(k\in\mathcal K^m\), where \(r_i^k\) is the (constant) value of \(r_i\) on cell \(k\).
Then \(\bs{\Psi}\), an approximation to \(\mathbb \Psi\), is given by the solution to the matrix-Ricatti equation 
\begin{align}\label{eqn:RiccatiPsi2}
    \bs D^{+-}
+ \bs \Psi   \bs D^{-+}\bs \Psi
+   \bs D^{++}\bs \Psi
+ \bs \Psi  \bs D^{--}
= \bs 0.
\end{align}
Due to the stochastic interpretation of the uniformisation and QBD-RAP schemes the approximations to \(\mathbb \Psi(s)\) have a stochastic interpretation as the first-return probabilities of a fluid queue driven by a CTMC and QBD-RAP, respectively (see \cite[Chapter~7]{p2019} and \cite{bgnp2021} for details on the latter). For the DG scheme the approximation of \(\mathbb \Psi(s)\) is not as well-understood. Here, we assume that it is given by the solution to (\ref{eqn:RiccatiPsi2}) and is a polynomial approximation. %\footnote{I believe that the resulting operator is a projection operator which, given an initial distribution, projects the distribution of \(X(\zeta_{Y}(\{0\}))\) on to a set of polynomial basis functions, much like the DG method itself.}

@@

To avoid the problems of oscillations we can sometimes employ a \emph{slope limiter} with the DG scheme which reduces the scheme to linear in the regions where oscillations are detected. We implemented two slope-limited DG schemes, the DG-lim scheme which takes a high order DG scheme and limits the solution as necessary, and the DG-lin-lim, which is a linear approximation on a finer grid designed to use approximately the same computational resources as the other schemes considered. The numerical experiments demonstrate a significant loss of accuracy in the approximation when a DG-lim scheme is used for discontinuous problems. We conclude that the DG-lim scheme is not a viable scheme for the discontinuous problems considered here. The numerical experiments suggest that the DG-lin-lim scheme can perform well, and is similar to the performance of the QBD-RAP scheme, in the presence of discontinuities. For highly discontinuous problems (i.e.~problems with point masses) the QBD-RAP scheme can outperform the DG-lin-lim scheme. %For the application of the schemes to fluid-fluid queues, there is no obvious way to apply the concept of a slope limiter. %When the slope limiter detects oscillations in the approximate solution, it reduces the DG scheme to linear in the region surrounding the oscillations, otherwise, the slope limiter leaves the solution unchanged. Thus, the slope limiter permits high-order approximations away from oscillations, while also removing oscillations.

@@

Lastly, we applied the DG, uniformisation and QBD-RAP schemes to two simple fluid-fluid queues in Section~\ref{sec: ffq num}. In the first fluid-fluid queue, which appears to have a smooth first return distribution, the DG scheme performs very well (however, the initial condition for the DG scheme displayed oscillations). Of the two positivity preserving schemes considered, the QBD-RAP scheme performs better than the uniformisation scheme. In the second example, which has a discontinuity in the first return distribution, the DG scheme produces the lowest errors, but exhibits oscillations in the solution. The QBD-RAP and uniformisation schemes do not produce oscillations and, of the two, the QBD-RAP scheme performs best.

As we mentioned in the introduction to this chapter, in Sections~\ref{sec: recon num}, \ref{sec: comp}, \ref{sec:stat}, and \ref{sec: ffq num} slope limiting is not used as it is not applicable: in these cases slope limiting is practically equivalent to post-processing the solution by detecting oscillations, then projecting oscillatory regions of the solution onto a basis of linear functions, and making some corrections to the solution if it happens to be negative.

Moreover, when computing approximations to the first-return operator of a fluid-fluid queue in Section~\ref{sec: ffq num}, the QBD-RAP and uniformisation schemes lead to an algorithm that is theoretically justified. In contrast, to-date there no theory surrounding the DG approximation to the same operator (let alone any positivity preserving variants of the DG scheme).

>>>>

We implement two slope limited DG schemes, the DG-lim and DG-lin-lim scheme, which we describe in more detail in the next section. The numerical experiments demonstrate that the DG-lim scheme does not lead to a viable approximation scheme in the presence of discontinuities, while the DG-lin-lim scheme results in a relatively performant approximation scheme for the discontinuous problems considered here. In general, the QBD-RAP scheme performs similarly to the DG-lin-lim scheme, while the uniformisation scheme is a viable approximation method, but produces larger errors than the QBD-RAP and DG-lin-lim scheme, for the examples considered here. When slope limiting is used, the resultant approximation is at best linear around discontinuities. Moreover, there is a computational cost in applying a limiter (or filter), which could be significant, particularly if many initial conditions are to be considered. In contrast, for the uniformisation and QBD-RAP, the approximate solution is guaranteed to produce positive probabilities without any post-processing or extra computation.

@@

%The concept of slope-limiting originates in the context of numerically integrating (typically with respect to a \emph{time} variable) a so-called \emph{semi-discrete system}, as we do in Section~\ref{sec: transient approx}. However, approximating the operator \(\mathbb\Psi\) of a fluid-fluid queue amounts to solving an implicit matrix-equation and there is no concept of numerical integration with respect to time. Hence, the concept of slope-limiting, it is not directly transferable. One could process the initial condition and post-process the solution with a limiter. However, there are possibly better choices for this post-processing, such as filtering, \citep[Section~5.6.1]{nodalDGBook}. One could also view the composition of the DG operator and the limiting operator as one, but the result is a non-linear operator which is of no use for simplifying the Riccati Equation which characterises \(\mathbb \Psi\). Considering there is no theory on the use of the DG scheme to approximate \(\mathbb \Psi\) and hence the resultant approximation is not rigorously-defined, the author does not know if the use of a limiter in this context is possible.

@@

We obtain approximations to the generator of the fluid queue \(\{(X(t),\varphi(t))\}\) via the DG, QBD-RAP and uniformisation schemes. We do not apply slope limiting in this case as slope limiting results in a non-linear operator which means slope limiting schemes cannot be used to approximate the operator-Riccati equation by a matrix-Riccati equation. All the discretisations use a cell width of \(\Delta=0.4\). The approximate generator matrix is then used to approximate the first-return operator \(\mathbb\Psi(s)\) as discussed in Section~\ref{sec: intro Psi}. As in Section~\ref{sec: intro Psi}, let \(\bs{B}^{m,n}\), \(m,n\in\{+,-,0\}\) be an approximation to \(\mathbb B^{m,n},\, m,n\in\{+,-,0\}\) of dimension \(p\), and recall that \(\bs{D}^{n,m},\,m,n\in\{+,-\}\) is given by 
\[\bs D^{m,n} = \bs{R}^m (\bs{B}^{m,n} + \bs{B}^{m,0}(\bs B^{0,0})^{-1}\bs{B}^{0,n}), \, m,n\in\{+,-\}, \]
where \(\bs R^m\) is a diagonal matrix with diagonal blocks \(1/r_i^k\bs{I}_p\), \(i\in\calS_k^m\), \(k\in\mathcal K^m\), where \(r_i^k\) is the (constant) value of \(r_i\) on cell \(k\).
Then \(\bs{\Psi}\), an approximation to \(\mathbb \Psi\), is given by the solution to the matrix-Ricatti equation 
\begin{align}\label{eqn:RiccatiPsi2}
    \bs D^{+-}
+ \bs \Psi   \bs D^{-+}\bs \Psi
+   \bs D^{++}\bs \Psi
+ \bs \Psi  \bs D^{--}
= \bs 0.
\end{align}
Due to the stochastic interpretation of the uniformisation and QBD-RAP schemes the approximations to \(\mathbb \Psi(s)\) have a stochastic interpretation as the first-return probabilities of a fluid queue driven by a CTMC and QBD-RAP, respectively (see \cite[Chapter~7]{p2019} and \cite{bgnp2021} for details on the latter). For the DG scheme the approximation of \(\mathbb \Psi(s)\) is not as well-understood. Here, we assume that it is given by the solution to (\ref{eqn:RiccatiPsi2}) and is a polynomial approximation. %\footnote{I believe that the resulting operator is a projection operator which, given an initial distribution, projects the distribution of \(X(\zeta_{Y}(\{0\}))\) on to a set of polynomial basis functions, much like the DG method itself.}

@@

To avoid the problems of oscillations we can sometimes employ a \emph{slope limiter} with the DG scheme which reduces the scheme to linear in the regions where oscillations are detected. We implemented two slope-limited DG schemes, the DG-lim scheme which takes a high order DG scheme and limits the solution as necessary, and the DG-lin-lim, which is a linear approximation on a finer grid designed to use approximately the same computational resources as the other schemes considered. The numerical experiments demonstrate a significant loss of accuracy in the approximation when a DG-lim scheme is used for discontinuous problems. We conclude that the DG-lim scheme is not a viable scheme for the discontinuous problems considered here. The numerical experiments suggest that the DG-lin-lim scheme can perform well, and is similar to the performance of the QBD-RAP scheme, in the presence of discontinuities. %For highly discontinuous problems (i.e.~problems with point masses) the QBD-RAP scheme can outperform the DG-lin-lim scheme. %For the application of the schemes to fluid-fluid queues, there is no obvious way to apply the concept of a slope limiter. %When the slope limiter detects oscillations in the approximate solution, it reduces the DG scheme to linear in the region surrounding the oscillations, otherwise, the slope limiter leaves the solution unchanged. Thus, the slope limiter permits high-order approximations away from oscillations, while also removing oscillations.

@@

Lastly, we applied the DG, uniformisation and QBD-RAP schemes to two simple fluid-fluid queues in Section~\ref{sec: ffq num}. In the first fluid-fluid queue, which appears to have a smooth first return distribution, the DG scheme performs very well (however, the initial condition for the DG scheme displayed oscillations). Of the two positivity preserving schemes considered, the QBD-RAP scheme performs better than the uniformisation scheme. In the second example, which has a discontinuity in the first return distribution, the DG scheme produces the lowest errors, but exhibits oscillations in the solution. The QBD-RAP and uniformisation schemes do not produce oscillations and, of the two, the QBD-RAP scheme performs best. Moreover, when computing approximations to the first-return operator of a fluid-fluid queue in Section~\ref{sec: ffq num}, the QBD-RAP and uniformisation schemes lead to an algorithm that is theoretically justified. In contrast, to-date there no theory surrounding the DG approximation to the same operator (let alone any positivity preserving variants of the DG scheme).

%As we mentioned in the introduction to this chapter, in Sections~\ref{sec: recon num}, \ref{sec:stat}, and \ref{sec: ffq num} slope limiting is not used as it is not applicable: in these cases slope limiting is practically equivalent to post-processing the solution by detecting oscillations, then projecting oscillatory regions of the solution onto a basis of linear functions, and making some corrections to the solution if it happens to be negative.

### 2022-09-19 chapter7/chapter.tex
finito

%!TEX root = ../thesis.tex
\chapter{Conclusion\label{ch: conclusion}} 
A fluid-fluid queue is a stochastic fluid queue, where the driving process is a fluid queue itself. Fluid queues provide a model for a single continuous performance measure of a system in the presence of a random environment. Fluid queues have found a wide variety of applications including risk processes, telecommunications, and environmental modelling, among others. Given the success of fluid queues it is plausible that the extension to fluid-fluid queues, which enable us to track two continuous performance measures of a system, will also find success. \cite{bo2014} provide an analysis of fluid-fluid queues and derive operator-analytic expressions for the first-return operator, and limiting distribution of a fluid-fluid queue.

Motivated by computation of the theoretical operators for stochastic fluid-fluid queues in \cite{bo2014}, this thesis investigates approximations method for the generator of stochastic fluid queues. Cell-based approximation methods are particularly useful as they allow flexible partitioning of the approximate operators, which is required in the analysis of fluid-fluid queues.

@@

Chapter~\ref{sec: numerics} investigates, via numerical experiments, the performance of the DG, QBD-RAP, and uniformisation schemes. In some contexts, we also implement two positivity preserving DG schemes which utilise a limiter. For smooth problems the DG scheme was superior, however, for discontinuous problems, the DG scheme exhibited oscillations and produced negative probability estimates. One of the slope limited DG schemes, in which a slope limiter is applied to a high-order DG scheme (which we title the DG-lim scheme), performed very poorly due to the limiter reducing the order of the scheme to linear near discontinuities. The QBD-RAP scheme performed similarly to the linear DG approximation on a finer grid with a slope limiter (the DG-lin-lim scheme). The QBD-RAP scheme performed slightly better on problems with point masses, and the DG-lin-lim scheme performed slightly better on the smoother problems. However, the QBD-RAP scheme constructs an approximation to the generator of the fluid queue which is linear and this is advantageous for the application to fluid-fluid queues. The uniformisation scheme performed better than the DG-lim scheme but worse than all the others.

In summary, this thesis investigated three main computational frameworks for the analysis of fluid queues and fluid-fluid queues, namely, the uniformisation, DG, and QBD-RAP schemes. The uniformisation scheme is from \cite{bo2013}. The DG method is well known, but its application to fluid and fluid-fluid queues has not been well-studied. The QBD-RAP scheme is a novel approach which overcomes some known issues of the DG scheme, and converges faster than the uniformisation scheme. Analysis of the QBD-RAP scheme proves that it is convergent. Numerical experiments demonstrate the effectiveness of the approximation schemes for various problems.

>>>>

%!TEX root = ../thesis.tex
\chapter{Conclusion\label{ch: conclusion}} 
A fluid-fluid queue is a stochastic fluid queue, where the driving process is a fluid queue itself. Fluid queues provide a model for a single continuous performance measure of a system in the presence of a random environment. Fluid queues have found a wide variety of applications including risk processes, telecommunications, and environmental modelling, among others. Given the success of fluid queues it is plausible that the extension to fluid-fluid queues, which enable us to track two continuous performance measures of a system, will also find success. \cite{bo2014} provide an analysis of fluid-fluid queues and derive operator-analytic expressions for the first-return operator, and limiting distribution of a fluid-fluid queue. Motivated by the computation of the theoretical operators for stochastic fluid-fluid queues in \cite{bo2014}, this thesis investigates approximations method for the generator of stochastic fluid queues. Cell-based approximation methods are particularly useful as they allow flexible partitioning of the approximate operators, which is required in the analysis of fluid-fluid queues.

@@

Chapter~\ref{sec: numerics} investigates, via numerical experiments, the performance of the DG, QBD-RAP, and uniformisation schemes. In some contexts, we also implement two positivity preserving DG schemes which utilise a limiter. For smooth problems the DG scheme was superior, however, for discontinuous problems, the DG scheme exhibited oscillations and produced negative probability estimates. One of the slope limited DG schemes, in which a slope limiter is applied to a high-order DG scheme (which we title the DG-lim scheme), performed very poorly due to the limiter reducing the order of the scheme to linear near discontinuities. The QBD-RAP scheme performed similarly to the linear DG approximation on a finer grid with a slope limiter (the DG-lin-lim scheme). %The QBD-RAP scheme performed slightly better on problems with point masses, and the DG-lin-lim scheme performed slightly better on the smoother problems. 
However, unlike the slope limited DG-lin-lim scheme, the QBD-RAP scheme constructs an approximation to the generator of the fluid queue which is linear and this is advantageous for the application to fluid-fluid queues. The uniformisation scheme performed better than the DG-lim scheme but worse than all the others.

In summary, this thesis investigated three main computational frameworks for the analysis of fluid queues and fluid-fluid queues, namely, the uniformisation, DG, and QBD-RAP schemes. The DG method is well known, but its application to fluid and fluid-fluid queues has not been well-studied. The QBD-RAP scheme is a novel approach which overcomes some known issues of the DG scheme, and converges faster than the uniformisation scheme. Analysis of the QBD-RAP scheme proves that it is convergent. Numerical experiments demonstrate the effectiveness of the approximation schemes for various problems.
